ML Estimation of Matrix t Distributions with Expectation Maximization

### 2016-12-11 paper.tex
Some work on the text
Improve wording and add a reference to the book.

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration of $S$ is the set of real symmetric positive definite
matrices with dimension $p$, $A$ is a $p\times p$ nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma function given in terms of the standard gamma function $\Gamma(\cdot)$ by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as $|\cdot|$. We also define the multivariate digamma function
\begin{equation}
\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

A random $p\times m$ matrix $T$ following a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$
has probability density
\begin{equation}
f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ and $\Omega$ are both real symmetric positive definite matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$ matrix indicates the offset, and $n$ is the number of degrees of freedom.

Following {[}book{]} we begin by deriving the probability density of matrix--$t$ distribution starting from a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$ with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale parameter $S$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has probability density
\begin{equation}
\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance
of the rows of $X$), $\Omega$ is a $m\times m$ positive definite
matrix (the covariance of the columns of $X$) and $M$ is a $p\times m$
matrix (the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the degrees of freedom and $\Gamma_{p}$ is the multivariate gamma
function defined in \ref{eq:multgammadef}. We will show that the
random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

@@

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In obtain to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve \ref{eq:firstlogpart} by substituting this result directly.

@@

For $\Omega_{t}$ we obtain:
\[
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\]
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated we have:
\[
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\[
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]

>>>>

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration of $S$ is the set of real symmetric
positive definite matrices with dimension $p$, $A$ is a $p\times p$
nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
function, given in terms of the standard gamma function $\Gamma(\cdot)$
by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as
$|\cdot|$. We also define the multivariate digamma function
%
\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

A random $p\times m$ matrix $T$ following a matrix--$t$ distribution
$T_{p,m}(M,\Sigma,\Omega,n)$ has the probability density
\begin{equation}
f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ and $\Omega$ are both real symmetric positive definite
matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$
matrix that indicates the location of the distribution, and $n$ is the
number of degrees of freedom.

Following \citet{gupta1999matrix}, we begin by deriving the
probability density of the matrix--$t$ distribution starting from
a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$
with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale
parameter $S$ of appropriate degrees of freedom $\nu$ (independent
from $X$), and marginalizing over $S$. Note that the row scale
parameter $S$ is a real symmetric positive definite matrix. We repeat
the results here because, along the way, we will utilize several
results that are useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has
the probability density %
\begin{equation}
	\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det
	S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance of
the rows of $X$), $\Omega$ is a $m\times m$ positive definite matrix
(the covariance of the columns of $X$) and $M$ is a $p\times m$ matrix
(the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the number of degrees of freedom and $\Gamma_{p}$ is the
multivariate gamma function defined in \ref{eq:multgammadef}. We will
show that the random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

@@

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In order to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.

@@

For $\Omega_{t}$ we obtain:
\[
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\]
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\[
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\[
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]
